---
title: "Tutorial"
output:
  github_document
always_allow_html: true
---

```{r include=FALSE}
library("vembedr")
```

## Novel regulatory effects

In this tutorial, we introduce and explain the regulatory effects predicted by DeepsmirUD and how users can use these data.

```{r echo=FALSE}
embed_youtube("OV5iSzzCkyY ") %>% 
  use_align("center") %>% 
  use_bs_responsive()
```

## Drug-cancer association

In this tutorial, we give the usage of the drug-cancer associations inferred by means of DeepsmirUD and a disease-related database.

```{r echo=FALSE}
embed_youtube("NybkuFPIB7I ") %>% 
  use_align("center") %>% 
  use_bs_responsive()
```

## Cancer-drug heatmap

In this tutorial, we show a cancer-drug heatmap to broaden the knowledge of the drug potential for a particular cancer disease.

```{r echo=FALSE}
embed_youtube("re4IpSYLWAQ ") %>% 
  use_align("center") %>% 
  use_bs_responsive()
```

**For more details, please check out the [paper](https://www.biorxiv.org/content/10.1101/2022.06.30.498304v2){target="_blank"} and [code](https://github.com/2003100127/deepsmirud){target="_blank"}.**
